Phylogenetics and the human microbiome.

### 2013-09-24 sbreview.tex
Still reorganizing and thinking.

Terminology: survey versus metagenome.
Definition of a single sample.
There are a number of non-sequence-based experimental techniques that have been crucial for the development of the field, such as manipulation of gnotobiotic mice, but these are beyond the scope of this paper.

@@

@@

Our modern knowledge of the microbial world is due to a large part derived from the work of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages.
Their work, and the scientists who followed them, used especially the 16S ribosomal gene.
This gene was chosen because...
They developed the first "universal" 16S PCR primers.
The great accomplishment of this is learning that archaea, although morphologically similar to bacteria, form their own separate lineage.

@@

As we will see, 16S is used frequently as a way of measuring diversity.
Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes may not be so much vertically inherited.
However, \citep{zaneveld2010ribosomal} have shown that there is a correlation between pairwise distances of 16S genes and genetic repertoire.
This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} using either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.

@@

Without a doubt, the greatest impact that phylogenetics has had on microbial ecology research is on changing our view of the taxonomic relationships between microorganisms.
In the absence of sequencing, microbiologist have to rely on microscopes and staining.
Lots of little things look alike.

The reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.

The most stunning example of this is the discovery of the archaea by Woese.

@@

Surprisingly little work on alpha diversity.

@@

@@

GEBA \cite{wu2009phylogeny}

@@

NJ

\cite{eddy1998profile}
Profile hidden Markov models.

\cite{li2006cdhit}
Li, W.  and Godzik, A.
\cite{edgar2010usearch}

\cite{price2010fasttree}
{FastTree 2--approximately maximum-likelihood trees for large alignments}

\cite{navlakha2009finding}
Finding biologically accurate clusterings in hierarchical tree decompositions using the variation of information

\cite{white2010alignment}
Alignment and clustering of phylogenetic markers-implications for microbial diversity studies

RNA copy number
\cite{morgan2010metagenomic}

@@

Alignment.

@@

Species delineation problem not explicitly considered. Clustering.

@@

There continues to be discussion about whether there is a tree of life.
Examples.
This can be a theoretical one.
What is the correct representation of the genetic ancestry of an organism that allows us to best predict.

@@

HGT.

\section{Discussion}
What we can expect next.
As described above, we can expect much more development in terms of specifics.
\cite{hooper2012interactions}
Except to the extent that it is useful to look at gene trees to understand function, phylogenenetics doesn't really have a role.

@@

There are limitations to what we can learn.
Genome doesn't tell all.
Host side is sometimes missing, but experiments are good.

Perhaps there is a role for trait evolution methods, or perhaps a simple method does prediction to within how good the data is.
There is the possibility that we don't have to think about individual organisms at all, but rather can think about the collection of genes and their metabolic network as a meta-organism (cite Elbo).
Question: to what extent is modeling cellular-level evolutionary history interesting?

>>>>

Terminology: survey versus metagenome.
Definition of a single sample.

@@

Gnotobiotic mice.
Human intervention.

@@

Our modern knowledge of the microbial world is due to a large part derived from the methods of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages.
Their work, and the scientists who followed them, used especially the 16S ribosomal gene.
This gene was chosen because...
They developed the first "universal" 16S PCR primers.
The great accomplishment of this is learning that archaea, although morphologically similar to bacteria, form their own separate lineage.

@@

@@

Phylogenetics has had a substantial impact on microbial ecology research by changing our view of the taxonomic relationships between microorganisms.
The most stunning example of this is the discovery of the archaea by Woese.
In the absence of sequencing, microbiologist have to rely on microscopes and staining.
Lots of little things look alike.

However, the reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.

@@

Despite this work, and the enthusiasm with which microbial ecologists have accepted between-community comparison, phylogenetic alpha diversity seems under-developed.
We have recently shown that partially-weighted abundance diversity measures do a good job of distinguishing between dysbiotic and "normal" states \cite{mccoy2013abundance}.
We have also looked at PD under rarefaction \cite{nipperess2013mean}.

@@

\subsection{Phylogeny and function}

As we will see below, 16S is frequently used as a way of measuring diversity.
Those accustomed to microbial genetics may think this surprising, because the genetic repertoire of microbes may not be so much vertically inherited.
However, \citep{zaneveld2010ribosomal} have shown that there is a correlation between pairwise distances of 16S genes and genetic repertoire.
This approach has recently been taken to its logical conclusion by trying to infer functional characteristics using discrete trait evolution models on 16S gene trees \citep{langille2013predictive} using either parsimony \citep{kluge1969quantitative} or likelihood \citep{pagel1994detecting} methods via the ape package \citep{paradis2004ape}.

@@

@@

\subsection{Alignment and tree inference}

@@

Alignment.
\cite{eddy1998profile}
Profile hidden Markov models.

PYNAST

NJ
\cite{price2010fasttree}
{FastTree 2--approximately maximum-likelihood trees for large alignments}

Species delineation problem not explicitly considered. Clustering.
\cite{li2006cdhit}
Li, W.  and Godzik, A.
\cite{edgar2010usearch}
\cite{navlakha2009finding} Finding biologically accurate clusterings in hierarchical tree decompositions using the variation of information
\cite{white2010alignment} Alignment and clustering of phylogenetic markers-implications for microbial diversity studies

@@

@@

There continues to be discussion about whether there is a tree of life.
Examples.
This can be a theoretical one, but here is a practical formalism: what is the correct representation of the genetic ancestry of an organism that allows us to best predict.

@@

The field has already benefited significantly from phylogeny using 16S.
Now we have more full-genome data.
We also have more full-genome methods.
What can we do with that?

HGT.
What about tree models of within-organism transfer?
Could we do tree shape statistics?

\section{Discussion}
What we can expect next.

@@

Perhaps there is a role for trait evolution methods, or perhaps a simple method does prediction to within how good the data is.
There is the possibility that we don't have to think about individual organisms at all, but rather can think about the collection of genes and their metabolic network as a meta-organism (cite Elbo).
Has limitation that you forget about cellular boundaries, which matter-- not a freely diffusing soup.
Question: to what extent is modeling cellular-level evolutionary history interesting?
As an extension microbes have to be close to one another-- requires population modeling.

There are limitations to what we can learn.
Genome doesn't tell all.
Host side is sometimes missing, but experiments are good.
As described above, we can expect much more development in terms of specifics.
\cite{hooper2012interactions}
